---
title: "Generative patterns in maelstrom"
vignette: >
  %\VignetteIndexEntry{Generative patterns in maelstrom}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(maelstrom)
```

## Euclidean rhythms

`euclid()` keeps only the hits of a Bjorklund distribution. So does the
mini-notation: `"bd(3,8)"` and `euclid(s("bd*8"), 3, 8)` are the same
rhythm.

```{r}
a <- vapply(query_arc(s("bd(3,8)"), 0, 1), hap_onset, numeric(1))
b <- vapply(query_arc(euclid(s("bd*8"), 3, 8), 0, 1), hap_onset, numeric(1))
a
identical(round(a, 9), round(b, 9))
```

Rotation shifts the pattern: `s("bd(3,8,2)")` starts two steps later.
Classic starting points: `bd(3,8)`, `hh(5,8)`, `cp(2,8)`.

## Polymeter with stack()

Layers do not have to agree on a cycle length. Stack a 3-step pattern
against a 4-step and a 5-step and they drift against each other,
locking back together every 60 steps:

```{r}
poly <- stack(s("bd*3"), s("hh*4"), s("cp*5"))
length(query_arc(poly, 0, 1))  # 3 + 4 + 5 events in one cycle
```

The `{a b, x y z}` mini-notation does the same thing inline:
`"{bd*3, hh*4}"`.

## Arpeggiate chords

`chord()` stacks notes into one event. `arp()` splits simultaneous
chord tones back into equal time slices, low to high:

```{r}
arped <- chord("c4", "e4", "g4") |> arp()
vapply(query_arc(arped, 0, 1), hap_onset, numeric(1))
```

One cycle, three slices. Drums and single notes pass through `arp()`
untouched, so you can arpeggiate a whole stack safely.

## Humanize and transpose

`humanize()` adds deterministic timing and velocity jitter, keyed by
event position, so the same pattern always humanizes the same way:

```{r}
tight <- s("bd*4")
loose <- s("bd*4") |> humanize(0.01, 0.15)
o1 <- vapply(query_arc(tight, 0, 1), hap_onset, numeric(1))
o2 <- vapply(query_arc(loose, 0, 1), hap_onset, numeric(1))
round(o2 - o1, 4)
```

`transpose()` shifts pitches by semitones:

```{r}
up <- note("c4 e4 g4") |> transpose(7)
vapply(query_arc(up, 0, 1), function(h) h$value$note, character(1))
```

## every: form over time

`every(n, fn)` applies `fn` once every `n` cycles. The classic use is
variation: a fill every 4 cycles, a reversal every 8.

```{r}
p <- s("bd*4") |> every(4, function(x) rev(x))
o3 <- vapply(query_arc(p, 0, 4), hap_onset, numeric(1))
length(o3)  # 16 kicks across 4 cycles, one cycle reversed
```

## Deterministic chance

`degrade()` drops events with a probability. It looks random, but it is
seeded per event: render the same pattern twice, get the same audio.

```{r}
thin <- s("hh*16") |> degrade(0.5)
n1 <- length(query_arc(thin, 0, 1))
n2 <- length(query_arc(thin, 0, 1))
c(n1, n2)
identical(
  render(thin, cycles = 1, cps = 1),
  render(thin, cycles = 1, cps = 1)
)
```

Same count, bit-identical audio. `sometimes()`, `often()`, `rarely()`,
and the `"bd?"` mini-notation all work the same way: chance you can
reproduce.

## A generative sketch

Euclidean drums, a polymeter hat layer, an arpeggiated chord that
changes every 4 cycles, and a humanized kick:

```{r}
drums <- stack(
  s("bd(3,8)") |> humanize(0.008, 0.1),
  s("~ sd(2,8) ~"),
  s("hh(5,8)") |> gain(0.5)
)
harmony <- chord("c3", "eb3", "g3", "bb3") |> arp() |> synth("pluck") |>
  every(4, function(x) transpose(x, -2)) |> gain(0.6)

sketch <- stack(drums, harmony)
w <- render(sketch, cycles = 4, cps = 1)
write_wav(file.path(tempdir(), "generative_sketch.wav"), w)
length(w)
```

Four seconds. Nothing in it was placed by hand except the chord and the
Euclidean parameters; the rest is structure. Change the numbers, get a
different piece with the same bones.
