---
title: "Getting started with maelstrom"
vignette: >
  %\VignetteIndexEntry{Getting started with maelstrom}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(maelstrom)
```

## What this is

maelstrom is a live-coding pattern engine in R. A pattern is a function
of time: give it a time span, it tells you which musical events happen
there. You write patterns in a compact mini-notation, shape them with
pipes, layer them with `stack()`, and render the result to WAV.

```{r}
s("bd sd hh cp")
```

That is a pattern. It does not make sound yet; it describes events. To
hear it, you render it.

## Mini-notation

`s()` takes sound names. `note()` takes pitches. `n()` takes numbers
(MIDI notes or ratios).

```{r}
s("bd sd hh cp")
note("c3 eb3 g3 bb3")
n("0 2 4 7")
```

The building blocks:

* `"~"` is a rest: `s("bd ~ sd ~")`
* `"*"` repeats: `"hh*8"` is eight hi-hats per cycle
* `"/"` slows down: `"bd/2"` plays every other cycle
* `"[ ]"` groups steps into one slot: `"bd [sd hh] cp"`
* `"bd(3,8)"` is a Euclidean rhythm: 3 hits spread over 8 steps

```{r}
vapply(query_arc(s("bd(3,8)"), 0, 1), hap_onset, numeric(1))
```

Three kick onsets, evenly spread. That is the whole idea: strings in,
structured time out.

## Pipes, pattern first

Every combinator takes the pattern first, so base R's `|>` reads like a
signal chain:

```{r}
s("bd*4") |> gain(0.8) |> lpf(1200)
```

Nothing has rendered yet. This is still a description: four kicks per
cycle, a bit quieter, lowpassed at 1200 Hz. You can keep chaining:
`pan()`, `echo()`, `swing()`, `drive()`, all the same shape.

## Layers with stack()

`stack()` plays patterns at the same time:

```{r}
kick  <- s("bd*4")
snare <- s("~ sd ~ sd")
hats  <- s("hh*8") |> swing(0.3) |> gain(0.5)

groove <- stack(kick, snare, hats)
```

Name your layers whatever you like. (The internal pattern constructor is
called `ms_pattern()`, precisely so a variable named `pattern` never
collides with it.)

## Rendering to WAV

`render()` turns cycles of a pattern into samples. `write_wav()` writes
them out.

```{r}
w <- render(groove, cycles = 2, cps = 1)
length(w)   # 2 cycles at 1 cycle/sec = 2 seconds of audio
max(abs(w)) # peaks are kept at or under 1
```

```{r}
wav <- file.path(tempdir(), "groove.wav")
write_wav(wav, w)
wav
```

`cps` is cycles per second: the tempo knob. `render_stereo()` gives you a
2-row stereo matrix instead, with per-event panning from `pan()`.

## A complete tiny track

Put it together: drums, a bass line, and a plucky lead with echo.

```{r}
bass <- note("c2 c2 g1 a1") |> synth("sine") |> lpf(400) |> gain(0.9)
lead <- note("c4 ~ eb4 ~") |> synth("pluck") |> gain(0.5) |>
  echo(n = 3, time = 0.25, fb = 0.4)

track <- stack(kick, snare, hats, bass, lead)
w <- render(track, cycles = 4, cps = 1)
write_wav(file.path(tempdir(), "tiny_track.wav"), w)
length(w)
```

Four seconds, five layers, one file. From here, the
sound-design vignette tours the voices and effects, and the generative
vignette covers Euclidean rhythms, polymeter, and deterministic chance.
