---
title: "Sound design in maelstrom"
vignette: >
  %\VignetteIndexEntry{Sound design in maelstrom}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(maelstrom)
```

## Two drum kits

maelstrom ships two drum kits. The default is fully synthesized; the
acoustic kit has a pitched kick, a tonal snare, and metallic hats.
Pick one per event with `kit()`:

```{r}
synth_kit    <- s("bd sd hh oh") |> kit("synth")
acoustic_kit <- s("bd sd hh oh") |> kit("acoustic")
```

Voices missing from the acoustic kit fall back to the synth kit, so you
can mix freely. The hat family (`hh`, `oh`, `shaker`) is leveled
internally to sit about 6-10 dB under the kick at the same gain; push
them with `gain()` if you want them hotter.

## The drum palette

Eleven synthesized voices, one per name:

```{r}
all_drums <- s("bd sd hh oh cp tom rim shaker ride crash conga")
w <- render(all_drums, cycles = 1, cps = 0.5)
length(w)  # one 2-second cycle, one hit per voice
```

`bd` kick, `sd` snare, `hh` closed hat, `oh` open hat, `cp` clap,
`tom`, `rim` rimshot, `shaker`, `ride`, `crash`, `conga`.

## Synth voices

`synth()` picks the oscillator for pitched notes. Five raw oscillators
plus four built voices:

```{r}
for (v in c("sine", "sawtooth", "square", "triangle", "noise")) {
  w <- render(note("a3") |> synth(v), cycles = 1, cps = 1)
  cat(v, "peak:", round(max(abs(w)), 3), "\n")
}
```

```{r}
for (v in c("pluck", "pad", "sub", "lead")) {
  w <- render(note("a3") |> synth(v), cycles = 1, cps = 1)
  cat(v, "peak:", round(max(abs(w)), 3), "\n")
}
```

`pluck` is a decaying Karplus-ish voice, `pad` is slow-attacking and
wide, `sub` is a deep sine stack, `lead` is a bright saw. Pair them with
`attack()` to shape the envelope:

```{r}
w <- render(note("c4") |> synth("pad") |> attack(0.2), cycles = 1, cps = 0.5)
max(abs(w))
```

## Filters, static and moving

`lpf()` takes a cutoff in Hz. It also takes a mini-notation string or a
pattern of numbers, and then the cutoff moves per event:

```{r}
static <- note("c3") |> synth("sawtooth") |> lpf(800)
moving <- s("bd*4") |> lpf(n("400 800 1600 3200"))
w <- render(moving, cycles = 1, cps = 1)
length(w)
```

Strings are parsed eagerly, so a typo like `lpf("oops")` fails when you
compose the pattern, not halfway through a render.

## Effects

Per-event effects, all pattern-first:

```{r}
s("bd*4") |> drive(2)                        # soft saturation
note("c3") |> synth("sawtooth") |> crush(4)  # bitcrush to 4 bits
note("c4") |> synth("pad") |> chorus(6)      # modulated short delay
s("bd ~ sd ~") |> echo(n = 3, time = 0.25, fb = 0.5)
```

`echo(n, time, fb)` adds `n` repeats, each `time` cycles later and
decaying by `fb`. `pan(x)` places events left (0) to right (1); it only
matters in a stereo render:

```{r}
w <- render_stereo(s("bd*4") |> pan(0.2), cycles = 1, cps = 1)
dim(w)  # 2 rows: left, right
```

`add_reverb()` works on rendered audio, mono or stereo. The wet signal
is computed in Rust, bit-exact with the pure-R reference, about 2000x
faster:

```{r}
wet <- add_reverb(w, amount = 0.3)
dim(wet)
```

Chain them like anything else. A dubby snare:

```{r}
dub_snare <- s("~ sd ~ sd") |> echo(n = 4, time = 0.375, fb = 0.5) |>
  pan(0.6) |> gain(0.8)
w <- render_stereo(stack(s("bd*4"), dub_snare), cycles = 2, cps = 1)
wet <- add_reverb(w, 0.25)
write_wav(file.path(tempdir(), "dub_snare.wav"), wet)
dim(wet)
```
